---
title: "case_studies"
output: html_document
date: "`r Sys.Date()`"
---

Let's take a look at other viruses from GMPD and Verena databases that fit the pattern we are interested in — infects multiple host species and has both contact and environmental transmission.

```{r}
library(here)
library(tidyverse)
library(magrittr)
library(rvest)
```


```{r}
gmpd <- read_csv(here("data/GMPD/GMPD_main.csv"))
gmpd_traits <- read_csv(here("data/GMPD/GMPD_parasite_traits.csv"))
clover <- read_csv(here("data/clover/CLOVER_1.0_Viruses_AssociationsFlatFile.csv"))
```

```{r}
gmpd_num_hosts <- gmpd %>% filter(ParType=="Virus") %>% dplyr::select(HostCorrectedName, ParasiteCorrectedName) %>% distinct() %>% group_by(ParasiteCorrectedName) %>% summarize(num_hosts = n())
gmpd_generalists <- gmpd_num_hosts %>% filter(num_hosts>1)

gmpd_close_nonclose <- gmpd_traits %>% filter(close+nonclose==2)

gmpd_select <- inner_join(gmpd_close_nonclose, gmpd_generalists, by = "ParasiteCorrectedName")
```


clover doesn't include data on transmission mode so will need to get those data from somewhere else if we want to use that

for now, we can take a look at the parasites that are positive from GMPD

Stealing this google scholar scraper from another project (doesn't always work)
```{r}
names <- tibble(species = gmpd_select$ParasiteCorrectedName)

names$results <- NA

#names$species <- tolower(names$species) 
names$species <- str_replace_all(string = names$species, pattern = " ", replacement = "+")


for (i in 1:nrow(names)){
  tmp <- pull(names[i,1])
  
  link <- str_replace(string="https://scholar.google.com/scholar?hl=en&as_sdt=0%2C11&q=REPLACEME&btnG=", pattern="REPLACEME", replacement=tmp)
  
  page <- read_html(link)
  
  summaries_css <- page %>% html_elements(css = ".gs_ab_mdw")
  
  text <- capture.output(print(summaries_css[2]))
  
  line <- stringr::str_extract(text, "About.+results")
  
  line_a <- str_replace(line, ",", "")
  
  line_b <- line_a[2]
  
  number <- as.numeric(str_extract(line_b, "[:digit:]+"))

  names$results[i] <- number
}

```


