---
title: "workshop"
author: "Daniel Suh"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
---


```{r, message = F}
library(here)

source(here("base","src.R"))

library(reshape2)
library(ggridges)
```


```{r, message = F}
data <- read_csv(here("data/weighted_prev_competence_111220.csv"))
tree <- ape::read.nexus(here("data/raw_data/asup_just_tree.txt"))
names <- read_csv(here("data/raw_data/species_names_ids.csv"))
```





## Variation in competence for each species

### comparing ranked means with and without log-transformation of raw data

```{r}
vl <- readRDS(here("processed_data","vl.rds"))

supp1_raw <- vl %>% arrange(mean) %>% 
  mutate(abb_name = factor(abb_name, levels = abb_name)) %>% 
  ggplot(.,aes(x=abb_name,y=mean)) +
  geom_point() + 
  xlab("Species") + ylab("Viral Load") + 
  geom_linerange(aes(ymin=mean-se,ymax=mean+se)) +
  geom_hline(yintercept = 1) + 
  labs(title = "Viral Loads") + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_y_continuous(breaks=seq(0,6,1))

supp1_log10 <- vl %>% arrange(mean) %>% 
  mutate(abb_name = factor(abb_name, levels = abb_name)) %>% 
  ggplot(.,aes(x=abb_name,y=log1p_mean)) +
  geom_point() + 
  xlab("Species") + ylab("log1p(Viral Load)") + 
  geom_linerange(aes(ymin=log1p_mean-log1p_se,ymax=log1p_mean+log1p_se)) +
  geom_hline(yintercept = 1) + 
  labs(title = "Viral Loads") + 
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  scale_y_continuous(breaks=seq(0,6,1))

supp1_raw
supp1_log10
```

# Incidence


```{r}
incidence <- data %>% group_by(WetAltID, Month.1) %>% summarize(incidence = sum(RV.Status))

incidence %>% ggplot(.,aes(x=Month.1, y=incidence)) + geom_col() + facet_grid(vars(WetAltID))
incidence %>% ggplot(.,aes(x=Month.1, y=incidence, group = 1)) + geom_point() + geom_line() + 
  facet_wrap(~WetAltID, nrow = 5, ncol = 4) + labs(title="Incidence over time") + theme(axis.text.x = element_text(angle = 75, vjust = 0.5, hjust=1))

data %>% filter(RV.Status==1) %>% 
  ggplot(., aes(x=Month.1,
                y=WetAltID,
                group=WetAltID,
                fill=as.factor(WetAltID))) + 
  geom_density_ridges_gradient(show.legend=F) +
  labs(title="# of RV cases per site over time")

incidence %>% ggplot(., aes(x=Month.1,
                            y=WetAltID,
                            group=WetAltID,
                            fill=as.factor(WetAltID),
                            height = incidence)) +
  geom_density_ridges_gradient(show.legend = F, stat = "identity")
```


# Prevalence

```{r}
comm_data <- readRDS(here("processed_data","comm_data.rds"))
comm_data %>% ungroup() %>% ggplot(.,aes(x=Month.1, y=Prevalence, group = 1)) + geom_point() + geom_line() + facet_wrap(~WetAltID, nrow = 5, ncol = 4) + labs(title="Prevalence over time")
```

# Abundance

```{r}
comm_data %>% ungroup() %>% ggplot(.,aes(x=Month.1, y=log10(size), group = 1)) + geom_point() + geom_line() + facet_wrap(~WetAltID, nrow = 5, ncol = 4) + labs(title="log10(Abundance) over time")
```

# Histograms

```{r}
comm_data %>% ggplot(.,aes(x=size)) + geom_histogram() + labs(title = "Untransformed abundance")
comm_data %>% ggplot(.,aes(x=log10(size))) + geom_histogram() + labs(title = "log10-transformed abundance")
comm_data %>% ggplot(.,aes(x=cc)) + geom_histogram() + labs(title = "Untransformed cc")
comm_data %>% ggplot(.,aes(x=log10(cc))) + geom_histogram() + labs(title = "log10-transformed cc")
comm_data %>% ggplot(.,aes(x=MeanWaterTempPredC)) + geom_histogram(binwidth=2) + labs(title = "Mean Water Temp")
```

# Effect Sizes

Standard deviations from the mean
This chunk attempts to characterize each wetland in terms of how exceptional they are compared to other wetlands in terms of abundance, cc, and temp
The idea is that if certain wetlands tend to be highly abundant with high community competence and lower temperatures, then we might expect them to be a perfect environment for ranavirus to thrive. 
```{r}
comm_data <- readRDS(here("processed_data","comm_data.rds"))
comm_data %>% mutate(mean = mean(size), sd = sd(size), var = mean(size)-size, size = size)
comm_effect <- comm_data %>% filter(MeanWaterTempPredC>0) %>% mutate(size_sds = (mean(size)-size)/sd(size),
                                    cc_sds = (mean(cc)-cc)/sd(cc),
                                    temp_sds = ((mean(MeanWaterTempPredC)-MeanWaterTempPredC)/sd(MeanWaterTempPredC))*-1) #multiply by -1 to make it opposite

effect_summ <- comm_effect %>% group_by(WetAltID) %>% summarize(size_over = sum(size_sds>0),
                                                                size_sum = sum(size_sds),
                                                                cc_over = sum(cc_sds>0),
                                                                cc_sum = sum(cc_sds),
                                                                temp_over = sum(temp_sds>0),
                                                                temp_sum = sum(temp_sds))
```


```{r}
library(plotly)
```

```{r}
plot_ly(data = effect_summ, x=effect_summ$temp_over, y=effect_summ$size_over, z=effect_summ$cc_over, 
        type="scatter3d", mode="markers", color=as.factor(effect_summ$WetAltID))

plot_ly(data = effect_summ, x=effect_summ$temp_sum, y=effect_summ$size_sum, z=effect_summ$cc_sum, 
        type="scatter3d", mode="markers", color=as.factor(effect_summ$WetAltID))
```


# Up to Peak

Redo analysis (effect size) with only the data from months before and including the peak. Rationale is that we are most interested in the features of a wetland that make it good or bad for an epidemic to occur. After an epidemic begins and peaks, the conditions of the wetland during the decline of the epidemic may not be as important since there are less susceptibles and parasite may be dying out. We'll see.


```{r}
comm_data_1 <- readRDS(here("processed_data","comm_data.rds"))
comm_data_1$month.n <- as.numeric(gsub("Month","",comm_data_1$Month.1))
comm_data_1 %<>% group_by(WetAltID) %>% mutate(max_mo = ifelse(Prevalence == max(Prevalence),month.n,0)) 
comm_data_1 %<>% group_by(WetAltID) %>% mutate(max_mo=max(max_mo))

comm_data_2 <- comm_data_1 %>% group_by(WetAltID) %>% filter(!month.n>max_mo) %>% ungroup()
```


```{r}
comm_data_2 %>% mutate(mean = mean(size), sd = sd(size), var = mean(size)-size, size = size)
comm_effect_2 <- comm_data_2 %>% filter(MeanWaterTempPredC>0) %>% mutate(size_sds = (mean(size)-size)/sd(size),
                                    cc_sds = (mean(cc)-cc)/sd(cc),
                                    temp_sds = ((mean(MeanWaterTempPredC)-MeanWaterTempPredC)/sd(MeanWaterTempPredC))*-1) #multiply by -1 to make it opposite

effect_summ_2 <- comm_effect_2 %>% group_by(WetAltID) %>% summarize(n = n(),
                                                                size_over = sum(size_sds>0),
                                                                size_sum = sum(size_sds),
                                                                cc_over = sum(cc_sds>0),
                                                                cc_sum = sum(cc_sds),
                                                                temp_over = sum(temp_sds>0),
                                                                temp_sum = sum(temp_sds))
effect_summ_2
```


```{r}
plot_ly(data = effect_summ_2, x=effect_summ_2$temp_over, y=effect_summ_2$size_over, z=effect_summ_2$cc_over, 
        type="scatter3d", mode="markers", color=as.factor(effect_summ_2$WetAltID))

plot_ly(data = effect_summ_2, x=effect_summ_2$temp_sum, y=effect_summ_2$size_sum, z=effect_summ_2$cc_sum, 
        type="scatter3d", mode="markers", color=as.factor(effect_summ_2$WetAltID))
```



Also, rerun dynamics without demography.


```{r}

```


# Phenology

Let's look at amphibian phenology from the data
 - when do amphibian larvae first appear
 - when are amphibian larvae no longer found (implies emergence)
 - abundance (average clutch size?)

I want each row to be one month in one wetland and have columns for each species abundance
```{r}
ph <- data %>% select(-c(X1, ID, Species, RV.Status, HostDensity, MinWaterTemp, MaxWaterTemp, MeanWaterT, CanopyCover, Stage, SQMean, LogSQMean, ArcSQMean, LogSQMean.1, Date.Collected, SVL, Zscore, Species.4:Species.42, log10_SQ, ln_SQ, log1p_SQ)) %>% filter(Stage2=="L") %>% distinct()
```

condensing this down was more difficult than expected
some things to ask Stacey about
 - some of this might be explained by measurements for one sampling period being taken on separate days
 - discrepancies exist for a single site_month
 - - 19_month4: CanopyCover
 - - 7_month6: HostDensity
 - lots of sites had the same MeanWaterTempPredC but different MeanWaterT
 - - these sites also had different MinWaterTemp and MaxWaterTemp

Also I need to reconsider stage with other analyses. There are about 90 adults. I am just going to filter them out for now and assume the values we have are for larvae only?
 - need to check on this. Do data include adults as well?
 
Data are now ready. ph only has one of every site-month combo
```{r}
ph_ab <- ph %>% pivot_longer(.,cols = c(AB2:AB42))
ph_ab %>% ggplot(.,aes(x=Month.1, y=value, color=name, group = name)) + geom_point() + geom_line() + facet_wrap(vars(WetAltID))
```

```{r}
ph_ra <- ph %>% pivot_longer(.,cols = c(RA2:RA42))
ph_ra %>% ggplot(.,aes(x=Month.1, y=value, color=name, group=name)) + geom_point() + geom_line() + facet_wrap(vars(WetAltID), ncol=1)
```

